Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-15 writeup_and_notes/male_dependent_drive.tex
pushing ybs minor edits

%FIgures 
{\bf Abstract:}
Because only one of four products of female meiosis is transmitted to offspring, oogenesis is particularly
	 susceptible to genomic conflicts arising when an allele gains in frequency at a cost to organismal fitness. 
In fact, many facets of oogenesis have been interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after fertilization - 
	potentially providing an opportunity for sperm to meddle with its outcome and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female meiosis when possible. 
These results are consistent with current knowledge of sperm-dependent meiotic drivers, 
	and suggest that the requirement of fertilization for the completion of meiosis may represent 
	yet another mechanism of female meiosis designed to prevent meiotic drive.
\newpage

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster, a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes \citep[pages
52-54][]{Burt2006} in these cases is challenging, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg  in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype. To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon})
\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$? \citep[see ][ for
        discussion]{Buckler1999}. }. %PUT THIS BACK IN
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors.

For comparative purposes, we present the standard drive model as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious in both sexes but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$) a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. The
        parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A, and where it can fix in the green
        and orange areasin Figure \ref{Invasion_space}B. When the allele is
        maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

>>>>

%FIgures 
{\bf Abstract:}
Because only one of four products of female meiosis is transmitted to offspring, o\"{o}genesis is particularly
	 susceptible to genomic conflicts arising when an allele gains in frequency at a cost to organismal fitness. 
In fact, many facets of oogenesis have been interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after fertilization - 
	potentially providing an opportunity for sperm to meddle with its outcome and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female meiosis when possible. 
These results are consistent with current knowledge of sperm-dependent meiotic drivers, 
	and suggest that the requirement of fertilization for the completion of meiosis may represent 
	yet another mechanism of female meiosis designed to prevent meiotic drive.
\newpage

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster, a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes \citep[pages
52-54 in][]{Burt2006} in these cases is challenging, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg  in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype (Model 1). To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon} and Model 2 in the Supplementary Worksheet)
\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$? \citep[see ][ for
        discussion]{Buckler1999}. }. %PUT THIS BACK IN
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors.

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \yb{\ref{Invasion_space}A}, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.
